Supp mat: time series paper

### 2018-02-19 writing/supp_mat_draft.tex
first update of short format version of the article

Here, we illustrate how we performed the subsampling and power
calculations for a specific population. We provide more detail here than
in the main manuscript. As an example, we examine a 35-year time series
of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global
Population Dynamics Database (NERC Centre for Population Biology 2010).
Simple linear regression indicates a significant decrease for this
population with an estimated slope coefficient of -0.0189581. We assume
that this significant increase over 35 years is in fact the ``true
trend``. In statistical jargon, the 35-year trend is an effect that is
actually present; we can reject the null hypothesis of no trend. We can
then use this as a benchmark to see if subsamples of the time series
also indicate a significant increase.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
time. The line is the best fit line from linear regresssion. (b)
Statistical power for different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

We then perform a subsampling routine to estimate the minimum time
required \(T_{min}\) (similar to Gerber, DeMaster, and Kareiva (1999)).
This is the same routine we used for the results in the main manuscript.

@@

\subsection{Predictors of minimum time
required}\label{predictors-of-minimum-time-required}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
\caption{Output of generalized linear model with a Poisson error
structure for predicting the minimum time required with explanatory
variables of the absolute value of the slope coefficient (or trend
strength), temporal autocorrelation, and variability in population
size.\label{fig:poisson_model}}
\end{figure}

In the main text, we explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
variance in population size. Here, we use a generalized linear model
framework with a Poisson error structure to determine explanatory
variables of the minimum time required. We use the same 878 populations
as in the main text. In figure \ref{fig:poisson_model}, we show a set of
residual plots for the regression. We then show the coefficient
estimates and levels of significance in table \ref{table:model_output}.

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine predictors of the
minimum time required for determing long-term population
trends.\label{table:model_output}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
0\tabularnewline
autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
\bottomrule
\end{longtable}

Our model with trend strength, autocorrelation, variance, and generation
length accounted for 72.45\% of the variation in the minimum time
required (Table \ref{table:model_output}). However, we also found trend
strength and variance to be strongly correlated with one another.
Therefore, we ran two additional models with either trend strength or
variance, but not both together. This resulted in lower explained
deviance (analogue to \(R^2\)) of 53.03\% and 45.03\%, respectively.

@@

\subsection{Minimum time required and biological
correlates}\label{minimum-time-required-and-biological-correlates}

In the main manuscript, we examined the minimum time required to detect
a significant trend in abundance over time using linear regression. As
detailed in the main manuscript the minimum time required was around 15,
but there was a wide distribution. Therefore, we were interested in
potential explanatory variables of the minimum time required. In the
main manuscript, we examined characteristics of the time series itself,
like variability, autocorrelation, and the trend in abundance over time.
Here, we combined our time series data with a data on life history
characteristics of amniotes from Myhrvold et al. (2015). There was life
history information available for 547 populations representing 315
different species, all of which were birds (Aves class).

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

We then correlated minimum time required for each population with its
corresponding life history characteristics. In figure
\ref{fig:biological_correlates} we examined minimum time required versus
generation length (years), litter size (n), adult body mass (grams),
maximum longevity (years), egg mass (grams), and incubation (days). None
of these variables had much explanatory power in accounting for the
variance in the minimum time required. We ran a generalized linear model
(with a Poisson error structure) and found that these six variables only
accounted for 6.87\% of the explained deviance in minimum time required.

@@

Minimum time required differed for different biological classes. For
instance, more time is required for species within the Actinopterygii
class compared to other species (Fig. \ref{fig:class}a). These
differences between biological classes can be explained by differences
in population variability, with species in the Actinopterygii class
experiencing larger inter-annual variability in population size (Fig.
\ref{fig:class}c).

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) long-term trend (estimated slope coefficient) by
species class, (c) interannual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Minimum time required to assess long-term trends in abundance
for values of statistical significance (\(\alpha\)) and power
(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
\end{figure}

@@

In the main text, we evaluated the minimum time required to determine
long-term trends in abundance via linear regression. This process
examined linear trends in abundance over time. Here we examine the
minimum time required to estimate long-term trends that are either
exponential growth or decay. We use the same methods as described in the
previous sections, but we take the \(log\) of population density, or
abundance.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in log(abundance) given power of
0.8.\label{fig:min_time_dist_log_pop}}
\end{figure}

@@

In the main text, we showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length (Bolker 2008; Johnson et al. 2015). The model in the
main text simulated linear population growth with only a slope
coefficient, y-intercept, and noise parameter required. This model is
purely phenomenological and does not include any species life history.
Here, we use the same routine as the main text, but simulate from a more
biologically-realistic population model. We use the model described in
White, Nagy, and Gruber (2014). The model is a stochastic,
age-structured population model that includes density-dependence for
lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Statistical power for different length of time series
simulations for a lemon shark population in Bimini,
Bahamas.\label{fig:shark_example}}
\end{figure}

Here, we parameterize the model for a situation where adult mortality
rate is high enough to cause a population decline. In the same way we
simulated models in the main text, we simulate this more
biologically-realistic model for different lengths of time. For each
length of time, we calculate the statistical power. Similar to results
with the simpler model, statistical power generally increases with
longer sampling time (Fig. \ref{fig:shark_example}). In this example,
the minimum time required (\(T_{min}\)) t0 obtain at least 0.8
statistical power, given a significance level of 0.05, is 27 years.

\section{Minimum time required to estimate geometric growth
rate}\label{minimum-time-required-to-estimate-geometric-growth-rate}

Instead of detecting a trend over time with linear regression, we could
also calculate the geometric growth rate of the population. In figure
\ref{fig:growth_rate}, we show how to calculate growth rates for
subsamples of a time series. First, we created subsamples of each
possible length from the full 35 year time series, as we did in the main
next. Next, we calculated the mean and standard deviation of growth
rates for each possible time series length (Fig.
\ref{fig:growth_rate}b). Lastly, we calculated the percent error
\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
between the mean of each time series length (observed) and the overall
population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
we show the percent error as a function of time series length. Here we
define the minimum time required as the minimum number of years to
achieve less than 20\% error.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
density over time, (b) mean and standard deviation of growth rate for
subsamples of entire time series, and (c) the percent error between mean
estimated growth rate and the true long-term growth rate. The vertical
bar denotes the minimum time required to estimate growth rate within
20\% error.\label{fig:growth_rate}}
\end{figure}

We applied the same calculations to 1032 population time series. We then
obtain a distribution of the minimum time required to measure the
``true" long-term growth rate (Fig. \ref{fig:min_time_growth_dist}). We
see a bimodal distribution with many populations required 30+ years to
estimate the long-term growth rate. The large number of short years
required is due to cases where the entire time series is consistently
increasing or decreasing at the same rate.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-11-1.pdf}
\caption{Histogram of the minimum time required in order to estimate the
long-term growth rate within 20\%
error.\label{fig:min_time_growth_dist}}
\end{figure}

@@

In the main text, we examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexibility (Wood 2006). GAMs are
models where a response variable depends on unknown smooth functions of
explanatory variables. GAMs, therefore, can identify relationships
between response variables and explanatory variables that are non-linear
and perhaps more complicated. The downside of GAMs is they typically
require more data and are also prone to overfitting.

Here, we conduct the same analyses in the main text, but instead
calculate the minimum time series required to detect trends over time
according to a GAM model. We hypothesized that GAMs should require less
time to detect a trend as they are more flexible than linear regression.
We provide an example that shows statistical power increases with more
time sampled (Fig. \ref{fig:gam_example}).

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
corresponding fitted GAM model in red and (b) statistical power as a
function of the number of years sampled. The horizontal ine at 0.8
indicates the minimum threshold for statistical power and the vertical
line denotes the minimum time required to achieve 0.8 statistical
power.\label{fig:gam_example}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-13-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance according to a GAM
model given statistical power of 0.8. The smoothing parameter was set to
3 for each population.\label{fig:min_time_dist_gam}}
\end{figure}

We then fit GAM models for 851 populations. We found a similar
distribution of minimum time required as in the main text for linear
regression (Fig. \ref{fig:min_time_dist_gam}). However, in line with our
hypothesis, the GAM models did result in a lower mean minimum time
required of 14.65 years compared to the results from the main text of
16.45 years.

@@

\hypertarget{ref-Myhrvold2015}{}
Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
life-history database to perform comparative analyses with birds,
mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.

>>>>

Here, I illustrate how we performed the subsampling and power
calculations for a specific population. As an example, I examine a
35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species
in the Global Population Dynamics Database (NERC Centre for Population
Biology 2010). Simple linear regression indicates a significant decrease
for this population with an estimated slope coefficient of -0.0271717. I
assume that this significant increase over 35 years is in fact the
``true trend``. In statistical jargon, the 35-year trend is an effect
that is actually present; we can reject the null hypothesis of no trend.
We can then use this as a benchmark to see if subsamples of the time
series also indicate a significant increase.

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
time. The line is the best fit line from linear regresssion. (b)
Statistical power for different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

I then perform a subsampling routine to estimate the minimum time
required \(T_{min}\) (similar to Gerber, DeMaster, and Kareiva (1999)).
This is the same routine I used for the results in the main manuscript.

@@

\subsection{Correlates of minimum time
required}\label{correlates-of-minimum-time-required}

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
\caption{Output of generalized linear model with a Poisson error
structure for predicting the minimum time required with explanatory
variables of the absolute value of the slope coefficient (or trend
strength), temporal autocorrelation, and coefficient of variation of
population size.\label{fig:poisson_model}}
\end{figure}

In the main text, I explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
coefficient of variation in population size. Here, I use a generalized
linear model framework with a Poisson error structure to determine
explanatory variables of the minimum time required. I use the same 822
populations as in the main text. In figure \ref{fig:poisson_model}, I
show a set of residual plots for the regression analysis. I then show
the coefficient estimates and levels of significance in table
\ref{table:model_summary_time_series_correlates}.

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine time series
characteristics as correlates of the minimum time required for
determining long-term population
trends.\label{table:model_summary_time_series_correlates}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 4.2658002 & 0.0369853 & 115.337838 &
0.0000000\tabularnewline
abs\_overall\_trend & -70.0413235 & 2.0027094 & -34.973283 &
0.0000000\tabularnewline
autocorrelation & 0.0601596 & 0.0550282 & 1.093249 &
0.2742843\tabularnewline
coefficient\_variation & -0.1928445 & 0.0387392 & -4.978022 &
0.0000006\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine life-history
trait correlates of the minimum time required for determining long-term
population
trends.\label{table:model_summary_biological_correlates}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 2.8131074 & 0.1297488 & 21.681181 &
0.0000000\tabularnewline
litter\_or\_clutch\_size\_n & 0.0473894 & 0.0058279 & 8.131439 &
0.0000000\tabularnewline
gen\_len & 0.0640424 & 0.0087382 & 7.328999 & 0.0000000\tabularnewline
adult\_body\_mass\_g & -0.0000809 & 0.0000177 & -4.563594 &
0.0000050\tabularnewline
maximum\_longevity\_y & -0.0074775 & 0.0019534 & -3.827907 &
0.0001292\tabularnewline
incubation\_d & -0.0127828 & 0.0031681 & -4.034864 &
0.0000546\tabularnewline
trophic2 & -0.2872678 & 0.1267799 & -2.265878 & 0.0234589\tabularnewline
trophic2.5 & -0.3604948 & 0.1260791 & -2.859274 &
0.0042461\tabularnewline
trophic3 & -0.2936987 & 0.1274698 & -2.304064 & 0.0212190\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine both time series
characteristics and life-history trait correlates of the minimum time
required for determining long-term population
trends.\label{table:model_summary_all_correlates}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 4.3236120 & 0.1361628 & 31.7532519 &
0.0000000\tabularnewline
abs(overall\_trend) & -70.8497548 & 2.8424505 & -24.9255898 &
0.0000000\tabularnewline
coefficient\_variation & -0.2089905 & 0.0531591 & -3.9314122 &
0.0000844\tabularnewline
autocorrelation & -0.0038698 & 0.0790400 & -0.0489597 &
0.9609515\tabularnewline
litter\_or\_clutch\_size\_n & 0.0008944 & 0.0058450 & 0.1530248 &
0.8783787\tabularnewline
gen\_len & 0.0087069 & 0.0088167 & 0.9875484 & 0.3233738\tabularnewline
adult\_body\_mass\_g & -0.0000214 & 0.0000174 & -1.2311406 &
0.2182703\tabularnewline
maximum\_longevity\_y & -0.0023131 & 0.0019647 & -1.1773197 &
0.2390679\tabularnewline
incubation\_d & -0.0004509 & 0.0031244 & -0.1443118 &
0.8852542\tabularnewline
trophic2 & 0.0302480 & 0.1273238 & 0.2375677 & 0.8122164\tabularnewline
trophic2.5 & -0.0101150 & 0.1263433 & -0.0800595 &
0.9361899\tabularnewline
trophic3 & -0.0049709 & 0.1276051 & -0.0389552 &
0.9689261\tabularnewline
\bottomrule
\end{longtable}

The model with trend strength, autocorrelation, and the coefficient of
variation in population size 75.13\% of the variation in the minimum
time required (Table \ref{table:model_summary_time_series_correlates}).

For a subset of populations, I examined life-history correlates as
potential explanatory variables for the minimum time required. Table
\ref{table:model_summary_biological_correlates} and
\ref{table:model_summary_all_correlates} summarize results for a model
with only life-history traits and an additional model with life-history
traits and time series characteristics combined, respectively. The model
with all life-history traits explained only 5.99\% of the variation in
the minimum time required (Table
\ref{table:model_summary_biological_correlates}). The model with all
life-history traits and time series characteristics explained 77.26\% of
the variation in the minimum time required (Table
\ref{table:model_summary_all_correlates}). When accounting for
characteristics of the time series, no life-history variables were
significant.

@@

@@

Minimum time required differed for different biological classes. For
instance, more time is required for species within the Actinopterygii
class compared to other species (Fig. \ref{fig:class}a). These
differences between biological classes can be explained by differences
in population variability, with species in the Actinopterygii class
experiencing larger inter-annual variability in population size (Fig.
\ref{fig:class}c). In addition, populations in the Actinopterygii class
generally had weaker trends in abundance (i.e.~smaller slope
coefficients) compared to the Aves and Mammalia populations.

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{(a) Minimum time required to estimate change in abundance by
biological class, (b) long-term trend (estimated slope coefficient) by
class, (c) coefficient of variation in population size by class, and (d)
temporal autocorrelation by class.\label{fig:class}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Minimum time required to assess long-term trends in abundance
for values of statistical significance (\(\alpha\)) and power
(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
\end{figure}

\subsection{Additional results for IUCN
analysis}\label{additional-results-for-iucn-analysis}

In the main text, I presented results detailing the difference between
the minimum time required to achieve a high level of statistical power
and the number of years suggested by IUCN based on generation length to
assess a species as vulnerable. There were deviations between the two
approaches. These deviations were mostly explained by the differences in
inter-annual population variability (coefficient of variation) and
generation time (Fig. \ref{fig:IUCN_correlates}).

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{The difference between minimum time estimates is the minimum
time required to achieve 0.8 statistical power versus the minimum time
required under IUCN criteria A2 to classify a species as vulnerable.
Each point represents a single population, all of which saw declines of
30\% or greater over a 10 year period. (a) Difference between minimum
time estimates versus the coefficient of variation in population size.
(b) Difference between minimum time estimates versus the generation
length in years.\label{fig:IUCN_correlates}}
\end{figure}

@@

In the main text, I evaluated the minimum time required to determine
long-term trends in abundance via linear regression. This process
examined linear trends in abundance over time. Here I examine the
minimum time required to estimate long-term trends that are either
exponential growth or decay. I use the same methods as described in the
previous sections, but we take the \(log\) of population density, or
abundance.

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-11-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in log(abundance) given power of
0.8.\label{fig:min_time_dist_log_pop}}
\end{figure}

@@

In the main text, I showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length (Bolker 2008; Johnson et al. 2015). The model in the
main text simulated linear population growth with only a slope
coefficient, y-intercept, and noise parameter required. This model is
purely phenomenological and does not include any species life-history
information. Here, I use the same routine as the main text, but simulate
from a more biologically-realistic population model. I use the model
described in White, Nagy, and Gruber (2014). The model is a stochastic,
age-structured population model that includes density-dependence for
lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{Statistical power for different length of time series
simulations for a lemon shark population in Bimini,
Bahamas.\label{fig:shark_example}}
\end{figure}

Here, I parameterize the model for a situation where adult mortality
rate is high enough to cause a population decline. As in the main text,
I simulated this more biologically-realistic model for different lengths
of time. For each length of time, I calculated the statistical power.
Similar to results with the simpler model, statistical power generally
increases with longer sampling time (Fig. \ref{fig:shark_example}). In
this example, the minimum time required (\(T_{min}\)) to obtain at least
0.8 statistical power, given a significance level of 0.05, is 27 years.

@@

In the main text, I examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexibility (Wood 2006). GAMs are
models where a response variable depends on unknown smooth functions of
explanatory variables. GAMs, therefore, can identify relationships
between response variables and explanatory variables that are non-linear
and perhaps more complicated. The downside of GAMs is they typically
require more data and are also prone to overfitting.

Here, I conduct the same analyses in the main text, but instead
calculate the minimum time series required to detect trends over time
according to a GAM model. I hypothesized that GAMs should require less
time to detect a trend as they are more flexible than linear regression.
I provide an example that shows statistical power increases with more
time sampled (Fig. \ref{fig:gam_example}).

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-15-1.pdf}
\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
corresponding fitted GAM model in red (with a smoothing parameter of 3)
and (b) statistical power as a function of the number of years sampled.
The horizontal ine at 0.8 indicates the minimum threshold for
statistical power and the vertical line denotes the minimum time
required to achieve 0.8 statistical power.\label{fig:gam_example}}
\end{figure}

\begin{figure}
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-16-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance according to a GAM
model given statistical power of 0.8. The smoothing parameter was set to
3 for each population.\label{fig:min_time_dist_gam}}
\end{figure}

We then fit GAM models for 799 populations. We found a similar
distribution of minimum time required as in the main text for linear
regression (Fig. \ref{fig:min_time_dist_gam}). However, in line with our
hypothesis, the GAM models did result in a lower mean minimum time
required of 14.12 years compared to the results from the main text of
15.91 years.

@@
